### 2020-06-09 chapters/sota.tex
Removal of the old poorly written and poorly categoryzed outlier detection methods

The authors define a time-series as \textit{"observations that have been recorded in an orderly fashion and which are correlated in time"}. Regarding the concept of \textit{outlier} the authors state that \textit{"outliers can be thought of as observations that do not follow the expected behavior"}. This definition is similar to the concept of outlier provided by Hawkins in 1980 where he states that an outlier is \textit{“an observation which deviates so much from other observations as to arouse suspicions that it was
generated by a different mechanism"} \cite{Hawkins-Outliers}. Furthermore, Blázquez-García et. al allude to the fact that outliers are often referred to in the literature as \textit{"anomalies, discordant observations, discords, exceptions, aberrations, surprises, peculiarities or contaminants"}. For the purpose of their review and this Thesis, all the previous terms refer to outliers.

@@

\subsection{A two window paradigm algorithm for change detection}
\label{subsec:2-window}

Kifer et al. propose an algorithm for change detection in data streams \cite{Kifer-Detecting-Change}. The authors propose that detecting change in data streams can be reduced to testing if two windows have different underlying distributions. Thus the change detection algorithm works on a two-window paradigm.

Figure \ref{fig:change-detection-2-windows} contains the pseudocode presented by the authors. The proposed method uses two windows, a reference \textit{W\textsubscript{1}} and a sliding one \textit{W\textsubscript{2}}. The authors use tuple-based windows of size \textit{k} and a true sliding window. The reference window \textit{W\textsubscript{2}} works as a \textit{baseline} and contains the first \textit{k} points of the stream that occurred after the last detected change.

The algorithm begins by filling both windows with the first \textit{k} tuples. Then it slides window \textit{W\textsubscript{2}} (by one tuple, since it is a true sliding window). It does this \textit{k} times and in between slides checks for change by applying the \textit{d} function, which measures the discrepancy between the two windows, \textit{W\textsubscript{1}} and \textit{W\textsubscript{2}}. Whenever this discrepancy is above a certain threshold $\alpha$, the algorithm reports a change in the data stream. Then it repeats the whole process, re-initializing both windows with the next \textit{k} items and proceeding as already described.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.8]{figures/2-windows-change-pseudocode.png}
      \caption{Two window method for data stream change detection}
      \label{fig:change-detection-2-windows}
    \end{center}
\end{figure}

This algorithm reduces change detection in data streams to testing whether two samples --- \textit{reference} and \textit{sliding} windows --- are generated by different distributions. The authors study methods for identifying differences in distribution between two samples and implement them as the \textit{d} function. Furthermore, they develop their own data structure and algorithm for efficient computation of \textit{d}.

%init - \textit{O(klog(k))}    
%runtime - \textit{O(klog(k)}

%k to initialize and when change is detected
%k * d
%d 
 %nlog(2k) to initialize tree and when change is detected
 %log(2k) to maintain the tree
\subsubsection*{Time complexity analysis}
Computing the value of \textit{d} can be done in \textit{O(log(k))} time per iteration. Hence, given \textit{k} iterations, the time complexity of this algorithm is \textit{O(k log(k))}.

\subsubsection*{Space complexity analysis}
The presented algorithm makes use of two \textit{k} sized windows and a custom data structure used for the efficient computation of \textit{d} of size \textit{k} as well. Since \textit{k} is constant, this method has constant space complexity.

\subsubsection*{Applicability to our Hypothesis}
In this Thesis, we want to detect and alert data pattern shifts so detecting change in data streams is directly applicable to our use case. However, our focus is on a lightweight solution and this algorithm has a linear space complexity, so it does not fit our scope as a possible solution.

%\subsection{Pattern Mining}

\subsection{Mining Frequent Patterns: the FP-Stream structure}
An alternative approach to detecting changes in data streams is to perform pattern mining. The idea behind using pattern mining algorithms being that if we can maintain a list of frequent patterns we would be able to detect when said patterns change --- \textit{i.e.} the list changes. Changes in the frequent mined patterns would be associated with a data pattern shift and immediately reported.

The authors of \cite{Giannella-Mining-Frequent-Patterns} developed a data structure named \textit{FP-Stream} --- that summarizes frequent data patterns -- and algorithms to build and incrementally maintain it.

We agree with the author's statement that it is \textit{"unrealistic to hold all streaming data in the limited main memory"}. With this in mind, they divide patterns into three categories: frequent, sub-frequent and infrequent patterns. The focus of their work is the mining and maintenance of frequent and sub-frequent patterns since the latter might become frequent later. Thus, infrequent patterns are discarded, using less memory.

The FP-Stream structure consists of a frequent pattern tree (FP-Tree \cite{Han-FP-tree}) with tilted-time windows in each of the nodes.

The authors are not very explicit on the definition of a tilted-time window. They claim we are \textit{"often interested in recent changes at a fine granularity, but long term changes at a coarse granularity"} and that tilted-time windows fit such use case. Figure \ref{fig:tilted-time-window} shows such a window and was withdrawn from the paper. In it, we see the 4 quarters of the last hour, then the last 24 hours and finally 31 days. They claim that \textit{"one can compute frequent itemsets in the last hour with the precision of quarter of an hour, the last day with the precision of hour, and so on, until the whole month"}.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=1]{figures/tilted-time-window.png}
      \caption{Tilted-time window}
      \label{fig:tilted-time-window}
    \end{center}
\end{figure}

An FP-Tree \cite{Han-FP-tree} as shown in Figure \ref{fig:fptree}, is a tree representation of frequent patterns. Each node in the frequent pattern tree represents a pattern and its frequency (\textit{support} column in the Figure), recorded in the node. The same tree construction algorithm from \cite{Han-FP-tree} is used in the FP-Stream algorithm.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=1]{figures/fptree.png}
      \caption{FP-Tree}
      \label{fig:fptree}
    \end{center}
\end{figure}

The authors propose using only one frequent pattern tree, where at each node, the frequency for each tilted-time window is maintained. Figure \ref{fig:fpstream} shows the FP-Stream structure, as an example of a frequent pattern tree with tilted-time windows embedded in each node.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.8]{figures/fp-stream.png}
      \caption{FP-Stream}
      \label{fig:fpstream}
    \end{center}
\end{figure}

The algorithm for constructing and maintaining the FP-Stream structure is given as a high-level list of instructions. The algorithm groups incoming streaming data into batches. Initialization is done only when the first batch is complete. As the transactions for the first batch arrived, the frequencies for all items are computed. Once all transactions for the first batch have arrived, the batch is scanned to create an FP-tree structure, pruning all infrequent items -- \textit{i.e.} below a certain frequency threshold. Finally, an FP-Stream structure is created by mining the frequent items from the FP-Tree.

The incremental update of the FP-Stream as data arrives is described on a very high level and does not add relevant information to understand time and space complexity without knowledge of the FP-Tree algorithm, on which it heavily relies. Further research will be done to perform a more thorough analysis of this algorithm, starting with the analysis of \cite{Han-FP-tree}.

\subsubsection*{Applicability to our Hypothesis}

No definitive conclusions can be made without the full time and space complexity analysis. However, the maintenance of a frequent pattern tree and tilted-time windows for each of the nodes might reveal to memory intensive, discarding this as a valid solution.

\subsection{Neighbor-Based Pattern Detection for Windows Over Streaming Data: the Extra-N algorithm}

The authors of \cite{Yang-Neighbor-Based-Pattern-Detection} propose a method for incremental detection of neighbor based patterns, namely density-based clusters and distance-based outliers, specifically for sliding windows scenarios. Figure \ref{fig:plot-cluster-outlier} is an example of the previous two mentioned neighbor based patterns. Formal definitions for both distance-based outliers \ref{fig:distance-outlier-def} and density-based clusters \ref{fig:density-cluster-def} are provided.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.8]{figures/density-distance-based.png}
      \caption[Clustering and outlier identification plot]{Two density-based clusters and one distance-based outlier determined by neighbour-based pattern detection algorithms}
      \label{fig:plot-cluster-outlier}
    \end{center}
\end{figure}

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.8]{figures/distance-based-outlier-def.png}
      \caption{Formal definition of distance-based outliers}
      \label{fig:distance-outlier-def}
    \end{center}
\end{figure}

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.8]{figures/density-based-cluster-def.png}
      \caption{Formal definition of density-based clusters}
      \label{fig:density-cluster-def}
    \end{center}
\end{figure}

Density-based clusters label data points as \textit{core points}, \textit{edge points} or \textit{noise}. These classifications are explained in \ref{fig:density-cluster-def}. Data points are classified based on the number of neighbors they have within a range $\theta\textsuperscript{range}$, measured versus a $\theta\textsuperscript{count}$. In brief, if a data point has more than $\theta\textsuperscript{count}$ neighbors within $\theta\textsuperscript{range}$ it is classified as a \textit{core point}. If it does not but it is a neighbor of a core point, then it is an \textit{edge point}. Otherwise, it is considered \textit{noise}.

The authors make use of the \textit{“predictability"} property of the expiration of existing objects. In other words, given a window with a fixed slide size, they predetermine the \textit{life-span} of any data point in the window on arrival --- \textit{i.e.} the future windows it will belong to. This leads to the notion of \textit{predicted views}. Given the objects in a current window, they predict the pattern structures that will persist in subsequent windows by considering the objects in the current window and abstract these predicted pattern structures into \textit{“predicted views"} of each future window. This technique allows the efficient maintenance of neighborship counts by pre-handling the eventual expiring effect of new data points.

In the paper, the authors first propose the Abstract-C and Abstract-M algorithms. Ultimately, they combine all the desirable properties of each algorithm into Extra-N.

\subsubsection{Abstract-C}
Abstract-C was the first proposed algorithm. The main idea is to maintain for each data point \textit{p\textsubscript{i}} the number of neighbors it has rather than the actual list of neighbors. The main advantage of this approach is the reduced memory footprint since only a count is stored --- a single integer --- instead of a list of pointers --- to all neighbors.

When a data point \textit{p\textsubscript{i}} expires, the count of each of \textit{p\textsubscript{i}}'s neighbors must be decremented by one --- since \textit{p\textsubscript{i}} no longer exists. However, we are unable to reach \textit{p\textsubscript{i}}'s neighbors since in Abstract-C only a count of neighbors is stored --- rather than pointers to each of the neighbors.

The solution to this problem is exploiting the previously presented \textit{predictability} property. The key idea is to predict the expiration of any data point \textit{p\textsubscript{i}} and pre-handle the impact it has on \textit{p\textsubscript{i}}'s neighbours. The authors introduce the concept of \textit{"lifetime neighbour counts"} or \textit{lt\_cnt}. The \textit{lt\_cnt} of a data point \textit{p\textsubscript{i}} corresponds to a sequence of \textit{predicted neighbour counts} --- \textit{i.e.} the number of \textit{predicted neighbours} \textit{p\textsubscript{i}} has in any future window where he exists. More precisely, each entry on \textit{pi.lt\_cnt} records the number of \textit{p\textsubscript{i}}'s current neighbours that are known to survive in the corresponding future window. Furthermore, the length of \textit{p\textsubscript{i}.lt\_cnt} is kept equal to \textit{p\textsubscript{i}.lifespan}, decreasing by one after each window slide, removing the left most entry. The authors provide the following example: consider a window \textit{W\textsubscript{i}} and a data point \textit{p\textsubscript{i}} with three neighbours, \textit{p\textsubscript{1}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}}. Applying the \textit{predictability} property makes it possible to compute the lifespan of all points. Assume \textit{p\textsubscript{1}} expires after \textit{W\textsubscript{i}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}} expire after \textit{W\textsubscript{i+1}} and \textit{p\textsubscript{i}} expires after \textit{W\textsubscript{i+2}}. This implies that at \textit{W\textsubscript{i}}, \textit{p\textsubscript{i}.lt\_cnt} = \{\textit{W\textsubscript{i}}: 3, \textit{W\textsubscript{i+1}}: 2, \textit{W\textsubscript{i+2}}: 0\}.

According to the definition given in \ref{fig:distance-outlier-def}, the lifetime neighbor counts (\textit{lt\_cnt}) carry enough information to compute distance-based outliers: for each data point \textit{p\textsubscript{i}} check if the first (current) element of \textit{p\textsubscript{i}.lt\_cnt} is less than  \textit{$\theta\textsuperscript{fra}$} $\times$ \textit{N} to decide whether a data point is an outlier or not, with \textit{N} as the number of data points. Similarly, the core objects for the density-based clusters can be computed by comparing the first element of \textit{p\textsubscript{i}.lt\_cnt} with \textit{$\theta$\textsuperscript{count}}. However, the \textit{lt\_cnt} array does not carry enough information to generate the density-based clusters. Despite being able  to find all core points in the window, we can not group them into clusters. In order to compute that, Abstract-C analyzes each core point and the surrounding $\theta\textsuperscript{range}$ area in the window to reconstruct the clusters.

In conclusion, Abstract-C achieves linear memory consumption in the number of data points in the window. However, since it reconstructs the clusters for each core object in the window, it has \textit{O(CP)} time complexity, with \textit{CP} as the total number of core points. Hence, its performance largely depends on  \textit{CP}, which can vary from 0 -- no core points --- to the total window size --- all core points.

\subsubsection{Abstract-M}
Abstract-M is proposed as an Abstract-C enhancement by introducing the notion of \textit{cluster memberships}, achieved by attributing a \textit{clusterID} to each data point.

The main challenge of this approach is discounting the effect of expired data points. The removal of any data point on any cluster may result in the total break of the cluster into multiple smaller ones, which may persist or degrade to noise. Furthermore, when a cluster is split, the newly formed clusters must be attributed with new \textit{clusterID}s --- \textit{i.e.} every single data point in those clusters must be relabeled with the new \textit{clusterID}.

Once again, the solution to handle expiring data points relies on the predictability property. Knowing the data points that exist in each window means we can also know the set of future clusters and points that belong. This is included in the \textit{predicted views} concept previously mentioned. In brief, the \textit{predicted clusters} are created using such predicted views for each of the future windows.

Addition of new data points may cause three changes to the clusters: \textit{birth} of a new cluster, \textit{expansion} of an existing one and \textit{merge} of multiple existing clusters. When adding a new data point, mark it with a new \textit{clusterID} in case of a \textit{birth} or in an \textit{expansion} case mark it with the \textit{clusterID} of the cluster it belongs to. Handling the \textit{merge} of multiple clusters is done by using a tree structure where two or more sibling \textit{clusterID}s are in fact part of a bigger cluster with the \textit{clusterID} stored in their parent. For example, if cluster \textit{ID1} and cluster \textit{ID2} are merged into a new cluster \textit{ID3} by the arrival of a new data point, they are inserted as children of the node containing \textit{ID3}.

New data points may join existing neighborhoods of existing data points and may promote them to core points by making the size of their neighborhood greater or equal to \textit{$\theta$\textsuperscript{count}}. Once such promotion happens, the promoted core point needs to communicate to its neighbors its new role, since noise in its neighborhood becomes edge points and is given a \textit{clusterID}. Since there are no connections from the promoted data point to its neighbors, Abstract-M analyzes each of the promoted core points and the $\theta\textsuperscript{range}$ area around them to find the points requiring state update.

In conclusion, Abstract-M remains linear in memory regarding the number of window data points. However, since the number of promoted core points tends to be smaller and is always less or equal to the number of core points, this is an improvement on Abstract-C.

\subsubsection{Extra-N}

Abstract-M still performs range queries (area with $\theta\textsuperscript{range}$ diameter) for each of the promoted core points. To improve upon this, Extra-N combines the neighborship maintenance mechanisms proposed in Abstract-C and Abstract-M.

The authors state that different information is required from different classes of data points (\textit{core}, \textit{edge} or \textit{noise}). For example, maintaining a list of points to neighbors is required for a non-core point while neighbor counts are sufficient for core points. More precisely, Extra-N marks each data point \textit{p\textsubscript{i}} with a \textit{clusterID} in all windows where \textit{p\textsubscript{i}} is predicted to be a core point while keeping the exact list of neighbors (pointers) of \textit{p\textsubscript{i}} for all windows where he is predicted to be noise or an edge point. This way, the Extra-N algorithm stores the amount of information needed to compute both distance-based outliers and density-based clusters according to their definitions, \ref{fig:distance-outlier-def} and \ref{fig:density-cluster-def} respectively.

\subsubsection*{Time complexity analysis}
Extra-N only performs range queries for incoming data points. This means that for each new data point \textit{p\textsubscript{i}}, Extra-N will search the area of diameter $\theta\textsuperscript{range}$ around it and update on it neighbors. In conclusion, the time complexity of Extra-N is \textit{O(n)} where \textit{n} is the number of new data points.

\subsubsection*{Space complexity analysis}
In summary, Extra-N keeps the memory consumption linear to the number of data points in the window without performing unnecessary range queries.

\subsubsection*{Applicability to our Hypothesis}
While a linear complexity fits many use cases, it does not fit ours since we aim to build a low memory footprint solution, ideally with constant space complexity. Hence, this algorithm will be of no use to us.

\section{Anomaly Detection}
Anomaly detection refers to the process of finding anomalies in data. An anomaly relates to a point in time where the behavior of the system does not match the expected one. Hence, under this definition, an anomaly does not necessarily imply a problem.

With the study of anomaly detection algorithms, we intend to search for algorithms that identify data pattern shifts as anomalies. If any of the approaches searched for is not computationally intensive it will be considered valid for the final solution.

\subsection{Real-Time Stream Anomaly Detection with Hierarchical Temporal Memory}

In \cite{Ahmad-HTM}, the authors present an anomaly detection technique based on an on-line algorithm called Hierarchical Temporal Memory (HTM). HTM has a few properties that are desirable for data stream anomaly detection scenarios. First, it is an unsupervised algorithm that requires no labels. This property comes in handy when monitoring a data stream in real-time, where labels for incoming data will likely be absent. Additionally, HTM adapts to noisy domains. In other words, random variations in the underlying distribution of incoming data that do not persist throughout time will be ignored. Furthermore, HTM detects both spatial and temporal anomalies --- \textit{i.e.} changes in magnitudes of data or unusual timing for particular patterns, respectively.

HTM is also an online learning algorithm, meaning that it adapts to changing statistics of the underlying data stream. In other words, if there is a sudden random increase in a specific data point attribute, the algorithm will produce an alert. However, if that initially thought to be a random spike becomes frequent then the algorithm adapts and assimilates this as "a new reality". Hence, it stops classifying it as an anomaly. Unfortunately, this property is not desirable in our system. Our stream monitoring system must report changes relatively to an initial static configuration and should not accept a new distribution as the new standard.

\subsubsection*{Applicability to our Hypothesis}
Online learning algorithms that adapt to data pattern shifts and eventually consider them regular patterns are not viable methods to accomplish our goal of building a monitoring system that alerts data pattern shifts continuously until they match the initial configuration again. Additionally, insights produced by machine learning models are not human-understandable, rendering the user unable to understand why an anomaly report was made. In conclusion, we do not deem HTM applicable to our solution.

\subsection{DeepAnT}

DeepAnT \cite{Munir-DeepAnT} is a deep learning-based approach for the detection of anomalies in time series data, that uses unlabeled data to capture and learn the data distribution of the stream used to forecast the normal behavior of the time series data. It is unsupervised - \textit{i.e.} it requires no labels - which is an advantage in a data streaming scenario where real-time decisions have to be made before labels are effectively-known. It is capable of detecting point anomalies in time series data with periodic and seasonal characteristics.

DeepAnT employs a Convolutional Neural Network (CNN) as its \textit{time series predictor} module. Before monitoring a data stream, the CNN must be trained. The model can be trained with a small data set while achieving good generalization capabilities due to the effective parameter sharing of the CNN. Additionally, the data set used to train the model thus not require labels and may up to 5\% of the data set may correspond to anomalous data.

\subsubsection*{Applicability to our Hypothesis}

Similar to the HTM algorithm presented, DeepAnT adapts to the underlying distribution of the data stream. In other words, if the data patterns shift from the initial configuration but stabilize, DeepAnT will accept it as it is. As mentioned, our goal is to build a monitoring system that alerts data pattern shifts continuously, until they match the initial configuration again. Furthermore, DeepAnT uses deep learning which is actually a subset of machine learning and just as inexplicable. In summary, we do not deem DeepAnT usable for our use case.

>>>>

The authors define a time-series as \textit{"observations that have been recorded in an orderly fashion and which are correlated in time"}. Regarding the concept of \textit{outlier} the authors state that \textit{"outliers can be thought of as observations that do not follow the expected behavior"}. This definition is similar to the concept of outlier provided by Hawkins in 1980 where he states that an outlier is \textit{“an observation which deviates so much from other observations as to arouse suspicions that it was
generated by a different mechanism"} \cite{Hawkins-Outliers}. Furthermore, Blázquez-García et. al allude to the fact that outliers are often referred to in the literature as \textit{"anomalies, discordant observations, discords, exceptions, aberrations, surprises, peculiarities or contaminants"} which for the purpose of their review and this Thesis all refer to outliers. Hence, outlier detection refers to the process of finding anomalies in data. An anomaly relates to a point in time where the behavior of the system does not match the expected one. Note that under this definition an anomaly does not necessarily imply a problem.

@@

Kifer et al. 2 window paradigm \\
ref window -> some data\\
sliding window -> current data\\
apply distance metric (like jsd) to both windows if above threshold alert and refill reference (adapts to new distro)\\
input data: uni or multi variate\\
method nature: univariate (doenst check correlation)\\
outlier type: subsequence outlier (window)\\

ExtraN\\
point outlier?\\
density based clusters\\
distance based outliers\\
predictability/forecast property based\\

HTM\\
point outlier, model based, prediction\\
black box ML algorithm\\
explain properties\\
use what authors say in their review\\

DeepAnT\\
point outlier, model based, prediction\\
black box DL algorithm\\
explain properties\\
use what authors say in their review\\
